When Proximity Isn't Enough: Network Segregation and SES Bias in Referrals

### 2025-08-12 writing/manuscript.tex
v1.2.1
todo: reviews from Sophie

\section*{Abstract}
The share of high-SES connections in one's network is a strong correlate of labor market income. We investigate whether SES biases in referral selection exacerbate differences high-SES connection shares. We conduct a lab-in-the-field experiment with 734 Colombian university students who make incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed bonus for referral recipients, we find that referrals go to high-performing peers with whom they take many courses together, regardless of incentives. While low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no biases towards their own and other groups, referring along their network shares. We find that network segregation, driven by program selection based on SES, limits cross-SES referral opportunities for even without an explicit SES bias. These suggest institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

\pagebreak
\section{Introduction}
Equally qualified individuals face different labor market outcomes based on their socioeconomic status \citep{stansbury2024class}. This persistent inequality undermines meritocratic ideals and represents a substantial barrier to economic mobility. A key driver of SES-based inequality in the labor market stems from differences in social capital.\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} Economic connectivity, defined as the share of high-SES connections among low-SES individuals, is an important facet of social capital because it correlates strongly with labor market income \citep{chetty2022na}. In this sense, a lack of social capital means lack of access to individuals with influential (higher paid) jobs and job opportunities. It implies having worse outcomes when using one's network to find jobs conditional on the capacity to leverage one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}

>>>>

\section*{Abstract}

The share of high-SES connections in one's network strongly correlates with labor market income. Understanding whether network-based inequalities stem from who people can meet versus whom they choose to connect with requires separating network composition from individual preferences. We achieve this separation through a lab-in-the-field experiment where 734 Colombian university students make incentivized referrals from their complete enrollment networks.

We conduct a lab-in-the-field experiment with 734 Colombian university students who make incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed bonus for referral recipients, we find that referrals go to high-performing peers with whom they take many courses together, regardless of the treatment. While low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no biases towards their own and other groups, referring along their network shares. We find that network segregation, driven by program selection based on SES, limits cross-SES referral opportunities for even without an explicit SES bias. These suggest institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

\pagebreak
\section{Introduction}
Equally qualified individuals face different labor market outcomes based on their socioeconomic status \citep{stansbury2024class}. This persistent inequality undermines meritocratic ideals and represents a substantial barrier to economic mobility. A key driver of SES-based inequality in the labor market stems from differences in social capital.\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} Economic connectedness, defined as the share of high-SES connections among low-SES individuals, is an important facet of social capital because it correlates strongly with labor market income \citep{chetty2022na}. In this sense, a lack of social capital means lack of access to individuals with influential (higher paid) jobs and job opportunities. It implies having worse outcomes when using one's network to find jobs conditional on the capacity to leverage one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}
